$p$-adic precision

### 2014-02-13 padicprec.tex
Working on introduction

\newcommand{\Z}{\mathbb Z}
\newcommand{\Zp}{\Z_p}
\newcommand{\Q}{\mathbb Q}
\newcommand{\Qp}{\Q_p}
\newcommand{\R}{\mathbb R}
\renewcommand{\O}{\mathcal O}
\newcommand{\XX}{\mathbf X}
\newcommand{\trans}{{}^{\text t}}

@@

During the last two decades, $p$-adic methods have become a tremendous tool in computer algebra, since it allows computation not feasible on finite fields, can control the growth of the size of the coefficients when dealing with rationnal numbers, or is needed when performing algorithm of $p$-adic nature.

Yet, $p$-adic algorithms are necessarily numeric or symbolic-numeric: one can not avoid working with finite precision, and thus, keeping track of loss in precision during the computation.

Here are some examples of algorithm that rely on $p$-adic computations:
\begin{itemize}
\item Bostan \cite{Bostan}, on how polynomials in $\mathbb{F}_p$ can be computed from theirs Newton series lifted in $\mathbb{Z}_p$;
\item Gaudry \cite{Gaudry}, on generation of genus $2$ curves via $p$-adic lifting;
\item Kedlaya \cite{Kedlaya}, on point counting on hyperelliptic curves using Monsky--Washnitzer cohomology;
\item Lercier \cite{Lercier}, on computing isogenies between elliptic curves using $p$-adic differential equations.
\end{itemize}

@@

The second source draws upon parallels between $\Qp$ and $\R$.  Both
occur as completions of $\Q$ and we represent elements in both in terms
of a set of distinguished rational numbers.
In $\R$, floating point arithmetic provides approximate operations $\oplus$ and $\odot$ on a
subset $S_{\infty,h} \subset \Z[\frac12]$ that model $+$ and $\cdot$ in $\R$ up to a given relative precision $h$:
\[
\lvert \frac{x \circledast y}{x \ast y} - 1 \rvert \le 2^{-h}
\]
for $\ast \in \{+, \cdot\}$ and all $x, y \in S_{\infty,h}$ with $x \ast y \ne 0$.  The $p$-adic analogue
defines floating point operations on $S_{p,h} \subset \Z[\frac1p]$ with
\[
\lvert \frac{x \circledast y}{x \ast y} - 1 \rvert_p \le p^{-h}.
\]
When using floating point arithmetic, elements are represented with a constant \emph{relative precision} $h$.

In the $p$-adic world, precision tracking using intervals is much more feasible.
Even a long sequence of operations may not sacrifice any precision.
The Gaussian distributions of measurement error over $\R$ have no direct analogue in
$\Qp$.  Instead, number theorists focus on provably determining
a result within a $p$-adic interval.  As a consequence, interval arithmetic is ubiquitous
in implementations of $p$-adic numbers.  % Add citations to Sage, Magma and PARI.

>>>>

\newcommand{\Z}{\mathbb Z}
\newcommand{\Zp}{\Z_p}
\newcommand{\Q}{\mathbb Q}
\newcommand{\Qp}{\Q_p}
\newcommand{\Fp}{\mathbb{F}_p}
\newcommand{\R}{\mathbb R}
\renewcommand{\O}{\mathcal O}
\newcommand{\XX}{\mathbf X}
\newcommand{\trans}{{}^{\text t}}

@@

The last two decades have seen a rise in the popularity of $p$-adic methods in
computational algebra.  For example,
\begin{itemize}
\item Bostan et al. \cite{Bostan} use Newton sums for polynomials over $\Zp$ to compute composed products for polynomials over $\Fp$;
\item Gaudry et al. \cite{Gaudry} use $p$-adic lifting methods to generate genus 2 CM hyperelliptic curves;
\item Kedlaya \cite{Kedlaya} uses $p$-adic cohomology to count points on hyperelliptic curves over finite fields;
\item Lercier and Survent \cite{Lercier} compute isogenies between elliptic curves using $p$-adic differential equations.
\end{itemize}
Like real numbers, most $p$-adic numbers cannot be represented exactly, but instead must be
stored with some finite precision.  In this paper we focus on methods for handling $p$-adic precision
that apply across many different algorithms.

The second source draws upon parallels between $\Qp$ and $\R$.  Both
occur as completions of $\Q$ and we represent elements of both in terms
of a set of distinguished rational numbers.
In $\R$, floating point arithmetic provides approximate operations $\oplus$ and $\odot$ on a
subset $S_{\infty,h} \subset \Z[\frac12]$ that model $+$ and $\cdot$ in $\R$ up to a given relative precision $h$:
\[
\lvert \frac{x \circledast y}{x \ast y} - 1 \rvert \le 2^{-h}
\]
for $\ast \in \{+, \cdot\}$ and all $x, y \in S_{\infty,h}$ with $x \ast y \ne 0$.  The $p$-adic analogue
defines floating point operations on $S_{p,h} \subset \Z[\frac1p]$ with
\[
\lvert \frac{x \circledast y}{x \ast y} - 1 \rvert_p \le p^{-h}.
\]
When using floating point arithmetic, elements are represented with a constant \emph{relative precision} $h$.

In the $p$-adic world, precision tracking using intervals is much more feasible.
Even a long sequence of operations may not sacrifice any precision.  Intervals
allow number theorists to provably determine a result modulo a given power of $p$,
and the Gaussian distributions of measurement error over $\R$ have no direct analogue
over $\Qp$ anyway.  As a consequence, interval arithmetic is ubiquitous
in implementations of $p$-adic numbers.  % Add citations to Sage, Magma and PARI.

@@
